---
title: "Data Science Draft"
author: "Shadae Lewis"
date: "3/27/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Notes:
1. Change heading names

Importing the data

Column names:
snoring - snoring range of the user (snoring rate) 
respiration - respiration rate
bodytemp - body temperature 
limb - limb movement rate 
bloodO - blood oxygen levels
eye - eye movement
sleephrs - number of hours of sleep
heart - heart rate 
stress - Stress Levels (0- low/normal, 1 – medium low, 2- medium, 3-medium high, 4 -high)

```{r}
sleep_stress = read.csv("./stress_sleep.csv",sep=",", header=TRUE)
sleep_stress
```


Playing with the data

```{r}
summary(sleep_stress)

#looks like there are no NA's or blank spaces in the data
#stress that we are investigating is not binary 

```

```{r}
#correlation
cor(sleep_stress)

pairs(sleep_stress)

#seems like there is a relationship with stress levels and all other variables
```

Decision Tree

```{r}
library(tree)
library(boot)

#convert stress to factor
sleep_stress$stress = as.character(sleep_stress$stress)

#Renaming the values
#(0- low/normal, 1 – medium low, 2- medium, 3-medium high, 4 -high)
sleep_stress$stress[sleep_stress$stress == '0'] = "low/normal"
sleep_stress$stress[sleep_stress$stress == '1'] = "medium low"
sleep_stress$stress[sleep_stress$stress == '2'] = "medium"
sleep_stress$stress[sleep_stress$stress == '3'] = "medium high"
sleep_stress$stress[sleep_stress$stress == '4'] = "high"

```


Before training and testing sets
```{r}
sleep_stress$stress = as.factor(sleep_stress$stress)

tree.sleep.stress = tree(stress~.,sleep_stress)
summary(tree.sleep.stress)

plot(tree.sleep.stress)
text(tree.sleep.stress,pretty = 0)

```


With training and testing sets
```{r}

set.seed(2)
sample =sample(nrow(sleep_stress), nrow(sleep_stress)*.8)
train = sleep_stress[sample,]
test = sleep_stress[-sample,]

#tree.model = tree(High ~ . -Sales, Carseats,subset = train)
#tree.pred = predict(tree.carseats,Carseats.test,type = "class")
#table(tree.pred,High.test) #random comment
```



